### 2019-11-26 ic-journal/part2.tex
text

Milner's book~\cite{Mil89} contains only two ``unique solutions of
equations'' theorems, one for strong equivalence, the other for
observational congruence.   There is however a
version  of the result for weak bisimilarity that shares a large portion of proof
steps with that case for observational congruence.  
As  weak bisimilarity 
 is not \hl{a} congruence, we have to be more restrictive on the syntax for
 the equation, using strong guardedness.
% to make correct statement of this
% theorem, we must slightly modify the concept of sequential expresses
% with further restrictions: no direct sums (or only guarded sums).
% The ``unique solution of equations'' theorem for weak equivalence
% requires more restrictions on the expression, namely, .
% Now we're ready to state and prove the following lemma:
\begin{lemma}[\texttt{WEAK_UNIQUE_SOLUTION_LEMMA}]
If the variable $X$ is guarded and sequential in $G$, and
$G\{P/X\}\overset{\alpha}{\rightarrow} P'$, then $P'$ takes the form
$H\{P/X\}$, for some expression $H$, and for any $Q$, we have
$G\{Q/X\}\overset{\alpha}{\rightarrow} H\{Q/X\}$. Moreover $H$ is
sequential, and if $\alpha = \tau$ then $H$ is also guarded.
\begin{alltt}
\HOLTokenTurnstile{} \HOLConst{SG} \HOLFreeVar{G} \HOLSymConst{\HOLTokenConj{}} \HOLConst{GSEQ} \HOLFreeVar{G} \HOLSymConst{\HOLTokenImp{}}
   \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{P} \HOLBoundVar{a} \ensuremath{\HOLBoundVar{P}\sp{\prime}}.
       \HOLFreeVar{G} \HOLBoundVar{P} \HOLTokenTransBegin\HOLBoundVar{a}\HOLTokenTransEnd \ensuremath{\HOLBoundVar{P}\sp{\prime}} \HOLSymConst{\HOLTokenImp{}}
       \HOLSymConst{\HOLTokenExists{}}\HOLBoundVar{H}. \HOLConst{GSEQ} \HOLBoundVar{H} \HOLSymConst{\HOLTokenConj{}} \ensuremath{(}\HOLBoundVar{a} \HOLSymConst{\ensuremath{=}} \HOLSymConst{\ensuremath{\tau}} \HOLSymConst{\HOLTokenImp{}} \HOLConst{SG} \HOLBoundVar{H}\ensuremath{)} \HOLSymConst{\HOLTokenConj{}} \ensuremath{\HOLBoundVar{P}\sp{\prime}} \HOLSymConst{\ensuremath{=}} \HOLBoundVar{H} \HOLBoundVar{P} \HOLSymConst{\HOLTokenConj{}} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{Q}. \HOLFreeVar{G} \HOLBoundVar{Q} \HOLTokenTransBegin\HOLBoundVar{a}\HOLTokenTransEnd \HOLBoundVar{H} \HOLBoundVar{Q}
\end{alltt}
\end{lemma}

@@

A delicate point in the formalisation of the results about unique solution of
contractions are the proof of Lemma~\ref{l:ruptocon} and lemmas alike;
in particular, there is
 an induction on the length of weak transitions. 
For this, rather than 
 introducing a refined form of weak transition relation
enriched with its length, 
we found it more elegant  to  work with traces
(a motivation for this is to set the ground for extensions of this
formalisation work to trace equivalence in place of bisimilarity).

@@

If the list of actions is empty, that means that there is no transition and hence,
if there is at most one visible action (i.e., a label) in the list of actions,
then the trace is also a weak transition. Here
we have to distinguish between two cases: no label and unique label (in
the list of actions). The definition of ``no
label'' in an action list is easy (here \texttt{MEM} tests if a given element is a member of a list):
\begin{alltt}
   \HOLConst{NO_LABEL} \HOLFreeVar{L} \HOLTokenDefEquality{} \HOLSymConst{\HOLTokenNeg{}}\HOLSymConst{\HOLTokenExists{}}\HOLBoundVar{l}. \HOLConst{MEM} \ensuremath{(}\HOLConst{label} \HOLBoundVar{l}\ensuremath{)} \HOLFreeVar{L}\hfill{[NO_LABEL_def]}
\end{alltt}

@@

The final relationship between traces and weak transitions is stated
and proved in the following theorem
(where the  variable $acts$ stands for
a list of actions); 
it says, a weak transition $P\overset{u}{\Rightarrow}P'$ is also a
trace $P\overset{acts}{\longrightarrow}P'$ with a
 non-empty action list $acts$, in which either there is no label (for $u = \tau$), or 
$u$ is the unique label (for $u \neq \tau$):
%\begin{lemma}
% A weak transition $P\overset{u}{\Rightarrow}P'$ is a just trace with non
% empty action list: 1) without any visible label, if $u = \tau$, or 2)
% $u$ is the unique label in the list, if $u \neq \tau$.
\begin{alltt}
\HOLTokenTurnstile{} \HOLFreeVar{P} \HOLTokenWeakTransBegin\HOLFreeVar{u}\HOLTokenWeakTransEnd \ensuremath{\HOLFreeVar{P}\sp{\prime}} \HOLSymConst{\HOLTokenEquiv{}}
   \HOLSymConst{\HOLTokenExists{}}\HOLBoundVar{acts}.
       \HOLConst{TRACE} \HOLFreeVar{P} \HOLBoundVar{acts} \ensuremath{\HOLFreeVar{P}\sp{\prime}} \HOLSymConst{\HOLTokenConj{}} \HOLSymConst{\HOLTokenNeg{}}\HOLConst{NULL} \HOLBoundVar{acts} \HOLSymConst{\HOLTokenConj{}}
       \HOLKeyword{if} \HOLFreeVar{u} \HOLSymConst{\ensuremath{=}} \HOLSymConst{\ensuremath{\tau}} \HOLKeyword{then} \HOLConst{NO_LABEL} \HOLBoundVar{acts} \HOLKeyword{else} \HOLConst{UNIQUE_LABEL} \HOLFreeVar{u} \HOLBoundVar{acts}\hfill{[WEAK_TRANS_AND_TRACE]}
\end{alltt}
%\end{lemma}

@@

Traces are actually used in the proof of above lemma via 
the following lemma (\texttt{unfolding_lemma4}):
\begin{alltt}
\HOLTokenTurnstile{} \HOLConst{GCONTEXT} \HOLFreeVar{C} \HOLSymConst{\HOLTokenConj{}} \HOLConst{WGS} \HOLFreeVar{E} \HOLSymConst{\HOLTokenConj{}} \HOLConst{TRACE} \ensuremath{(}\ensuremath{(}\HOLFreeVar{C} \HOLSymConst{\HOLTokenCompose} \HOLConst{FUNPOW} \HOLFreeVar{E} \HOLFreeVar{n}\ensuremath{)} \HOLFreeVar{P}\ensuremath{)} \HOLFreeVar{xs} \ensuremath{\HOLFreeVar{P}\sp{\prime}} \HOLSymConst{\HOLTokenConj{}} \HOLConst{LENGTH} \HOLFreeVar{xs} \HOLSymConst{\HOLTokenLeq{}} \HOLFreeVar{n} \HOLSymConst{\HOLTokenImp{}}
   \HOLSymConst{\HOLTokenExists{}}\ensuremath{\HOLBoundVar{C}\sp{\prime}}. \HOLConst{GCONTEXT} \ensuremath{\HOLBoundVar{C}\sp{\prime}} \HOLSymConst{\HOLTokenConj{}} \ensuremath{\HOLFreeVar{P}\sp{\prime}} \HOLSymConst{\ensuremath{=}} \ensuremath{\HOLBoundVar{C}\sp{\prime}} \HOLFreeVar{P} \HOLSymConst{\HOLTokenConj{}} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{Q}. \HOLConst{TRACE} \ensuremath{(}\ensuremath{(}\HOLFreeVar{C} \HOLSymConst{\HOLTokenCompose} \HOLConst{FUNPOW} \HOLFreeVar{E} \HOLFreeVar{n}\ensuremath{)} \HOLBoundVar{Q}\ensuremath{)} \HOLFreeVar{xs} \ensuremath{(}\ensuremath{\HOLBoundVar{C}\sp{\prime}} \HOLBoundVar{Q}\ensuremath{)}
\end{alltt}
which roughly says, for any context $C$ and weakly-guarded context
$E$, if $C [\, E^n[P]\,] \overset{xs}{\Longrightarrow} P'$ and the length
of actions $xs \leqslant n$, then \hl{$P'$} has the form of $C'[P]$.
Traces are used \hl{in} reasoning about the number of intermediate actions in weak
transitions. For instance, from Def.~\ref{d:BisCon}, it is easy
to see that, a weak transition either becomes shorter
or remains the same when moving between $\mcontrBIS$-related processes.
This property is essential in the proof of
Lemma~\ref{l:uptocon}. We show only one such lemma, for the case of
non-$\tau$ weak transitions passing into $\mcontrBIS$ (from left to right):
\begin{alltt}
\HOLTokenTurnstile{} \HOLFreeVar{P} \HOLSymConst{\HOLTokenContracts{}} \HOLFreeVar{Q} \HOLSymConst{\HOLTokenImp{}}
   \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{xs} \HOLBoundVar{l} \ensuremath{\HOLBoundVar{P}\sp{\prime}}.
       \HOLConst{TRACE} \HOLFreeVar{P} \HOLBoundVar{xs} \ensuremath{\HOLBoundVar{P}\sp{\prime}} \HOLSymConst{\HOLTokenConj{}} \HOLConst{UNIQUE_LABEL} \ensuremath{(}\HOLConst{label} \HOLBoundVar{l}\ensuremath{)} \HOLBoundVar{xs} \HOLSymConst{\HOLTokenImp{}}
       \HOLSymConst{\HOLTokenExists{}}\ensuremath{\HOLBoundVar{xs}\sp{\prime}} \ensuremath{\HOLBoundVar{Q}\sp{\prime}}.
           \HOLConst{TRACE} \HOLFreeVar{Q} \ensuremath{\HOLBoundVar{xs}\sp{\prime}} \ensuremath{\HOLBoundVar{Q}\sp{\prime}} \HOLSymConst{\HOLTokenConj{}} \HOLFreeVar{P} \HOLSymConst{\HOLTokenContracts{}} \HOLFreeVar{Q} \HOLSymConst{\HOLTokenConj{}} \HOLConst{LENGTH} \ensuremath{\HOLBoundVar{xs}\sp{\prime}} \HOLSymConst{\HOLTokenLeq{}} \HOLConst{LENGTH} \HOLBoundVar{xs} \HOLSymConst{\HOLTokenConj{}}
           \HOLConst{UNIQUE_LABEL} \ensuremath{(}\HOLConst{label} \HOLBoundVar{l}\ensuremath{)} \ensuremath{\HOLBoundVar{xs}\sp{\prime}}\hfill{[contracts_AND_TRACE_label]}
\end{alltt}

>>>>

Milner's book~\cite{Mil89} contains only two ``unique solutions of
equations'' theorems, one for strong equivalence, the other for
observational congruence.   There is however a
version  of the result for weak bisimilarity that shares a large portion of proof
steps with the proof for observational congruence.  
As  weak bisimilarity 
 is not \hl{a} congruence, we have to be more restrictive on the syntax for
 the equation, using strong guardedness.
% to make correct statement of this
% theorem, we must slightly modify the concept of sequential expresses
% with further restrictions: no direct sums (or only guarded sums).
% The ``unique solution of equations'' theorem for weak equivalence
% requires more restrictions on the expression, namely, .
% Now we're ready to state and prove the following lemma:
\begin{lemma}[\texttt{WEAK_UNIQUE_SOLUTION_LEMMA}]
If the variable $X$ is guarded and sequential in $G$, and
$G\{P/X\}\overset{\alpha}{\rightarrow} P'$, then $P'$ takes the form
$H\{P/X\}$, for some expression $H$, and for any $Q$, we have
$G\{Q/X\}\overset{\alpha}{\rightarrow} H\{Q/X\}$. Moreover $H$ is
sequential, and if $\alpha = \tau$, then $H$ is also guarded.
\begin{alltt}
\HOLTokenTurnstile{} \HOLConst{SG} \HOLFreeVar{G} \HOLSymConst{\HOLTokenConj{}} \HOLConst{GSEQ} \HOLFreeVar{G} \HOLSymConst{\HOLTokenImp{}}
   \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{P} \HOLBoundVar{a} \ensuremath{\HOLBoundVar{P}\sp{\prime}}.
       \HOLFreeVar{G} \HOLBoundVar{P} \HOLTokenTransBegin\HOLBoundVar{a}\HOLTokenTransEnd \ensuremath{\HOLBoundVar{P}\sp{\prime}} \HOLSymConst{\HOLTokenImp{}}
       \HOLSymConst{\HOLTokenExists{}}\HOLBoundVar{H}. \HOLConst{GSEQ} \HOLBoundVar{H} \HOLSymConst{\HOLTokenConj{}} \ensuremath{(}\HOLBoundVar{a} \HOLSymConst{\ensuremath{=}} \HOLSymConst{\ensuremath{\tau}} \HOLSymConst{\HOLTokenImp{}} \HOLConst{SG} \HOLBoundVar{H}\ensuremath{)} \HOLSymConst{\HOLTokenConj{}} \ensuremath{\HOLBoundVar{P}\sp{\prime}} \HOLSymConst{\ensuremath{=}} \HOLBoundVar{H} \HOLBoundVar{P} \HOLSymConst{\HOLTokenConj{}} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{Q}. \HOLFreeVar{G} \HOLBoundVar{Q} \HOLTokenTransBegin\HOLBoundVar{a}\HOLTokenTransEnd \HOLBoundVar{H} \HOLBoundVar{Q}
\end{alltt}
\end{lemma}

@@

A delicate point in the formalisation of the results about unique solution of
contractions are the proof of Lemma~\ref{l:ruptocon} and lemmas alike;
in particular, there is
 an induction on the length of weak transitions. 
For this, rather than 
 introducing a refined form of weak transition relation
enriched with its length, 
we found it more elegant  to  work with traces
(a further motivation  is to set the ground for extensions of this
formalisation work to trace equivalence in place of bisimilarity).

@@

If the list of actions is empty,  there is no transition and hence,
if there is at most one visible action (i.e., a label) in the list of actions,
then the trace is also a weak transition. Here
we have to distinguish between two cases: no label and unique label (in
the list of actions). The definition of ``no
label'' in an action list is easy (here \texttt{MEM} tests if a given element is a member of a list):
\begin{alltt}
   \HOLConst{NO_LABEL} \HOLFreeVar{L} \HOLTokenDefEquality{} \HOLSymConst{\HOLTokenNeg{}}\HOLSymConst{\HOLTokenExists{}}\HOLBoundVar{l}. \HOLConst{MEM} \ensuremath{(}\HOLConst{label} \HOLBoundVar{l}\ensuremath{)} \HOLFreeVar{L}\hfill{[NO_LABEL_def]}
\end{alltt}

@@

The final relationship between traces and weak transitions is stated
and proved in the following theorem
(where the  variable $acts$ stands for
a list of actions); 
it says that a weak transition $P\overset{u}{\Rightarrow}P'$ is also a
trace $P\overset{acts}{\longrightarrow}P'$ with a
 non-empty action list $acts$, in which either there is no label (for $u = \tau$), or 
$u$ is the unique label (for $u \neq \tau$):
%\begin{lemma}
% A weak transition $P\overset{u}{\Rightarrow}P'$ is a just trace with non
% empty action list: 1) without any visible label, if $u = \tau$, or 2)
% $u$ is the unique label in the list, if $u \neq \tau$.
\begin{alltt}
\HOLTokenTurnstile{} \HOLFreeVar{P} \HOLTokenWeakTransBegin\HOLFreeVar{u}\HOLTokenWeakTransEnd \ensuremath{\HOLFreeVar{P}\sp{\prime}} \HOLSymConst{\HOLTokenEquiv{}}
   \HOLSymConst{\HOLTokenExists{}}\HOLBoundVar{acts}.
       \HOLConst{TRACE} \HOLFreeVar{P} \HOLBoundVar{acts} \ensuremath{\HOLFreeVar{P}\sp{\prime}} \HOLSymConst{\HOLTokenConj{}} \HOLSymConst{\HOLTokenNeg{}}\HOLConst{NULL} \HOLBoundVar{acts} \HOLSymConst{\HOLTokenConj{}}
       \HOLKeyword{if} \HOLFreeVar{u} \HOLSymConst{\ensuremath{=}} \HOLSymConst{\ensuremath{\tau}} \HOLKeyword{then} \HOLConst{NO_LABEL} \HOLBoundVar{acts} \HOLKeyword{else} \HOLConst{UNIQUE_LABEL} \HOLFreeVar{u} \HOLBoundVar{acts}\hfill{[WEAK_TRANS_AND_TRACE]}
\end{alltt}
%\end{lemma}

@@

Traces are actually used in the proof of above lemma via 
the following lemma (\texttt{unfolding_lemma4}):
\begin{alltt}
\HOLTokenTurnstile{} \HOLConst{GCONTEXT} \HOLFreeVar{C} \HOLSymConst{\HOLTokenConj{}} \HOLConst{WGS} \HOLFreeVar{E} \HOLSymConst{\HOLTokenConj{}} \HOLConst{TRACE} \ensuremath{(}\ensuremath{(}\HOLFreeVar{C} \HOLSymConst{\HOLTokenCompose} \HOLConst{FUNPOW} \HOLFreeVar{E} \HOLFreeVar{n}\ensuremath{)} \HOLFreeVar{P}\ensuremath{)} \HOLFreeVar{xs} \ensuremath{\HOLFreeVar{P}\sp{\prime}} \HOLSymConst{\HOLTokenConj{}} \HOLConst{LENGTH} \HOLFreeVar{xs} \HOLSymConst{\HOLTokenLeq{}} \HOLFreeVar{n} \HOLSymConst{\HOLTokenImp{}}
   \HOLSymConst{\HOLTokenExists{}}\ensuremath{\HOLBoundVar{C}\sp{\prime}}. \HOLConst{GCONTEXT} \ensuremath{\HOLBoundVar{C}\sp{\prime}} \HOLSymConst{\HOLTokenConj{}} \ensuremath{\HOLFreeVar{P}\sp{\prime}} \HOLSymConst{\ensuremath{=}} \ensuremath{\HOLBoundVar{C}\sp{\prime}} \HOLFreeVar{P} \HOLSymConst{\HOLTokenConj{}} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{Q}. \HOLConst{TRACE} \ensuremath{(}\ensuremath{(}\HOLFreeVar{C} \HOLSymConst{\HOLTokenCompose} \HOLConst{FUNPOW} \HOLFreeVar{E} \HOLFreeVar{n}\ensuremath{)} \HOLBoundVar{Q}\ensuremath{)} \HOLFreeVar{xs} \ensuremath{(}\ensuremath{\HOLBoundVar{C}\sp{\prime}} \HOLBoundVar{Q}\ensuremath{)}
\end{alltt}
which roughly says that for any context $C$ and weakly-guarded context
$E$, if $C [\, E^n[P]\,] \overset{xs}{\Longrightarrow} P'$ and the length
of actions $xs \leqslant n$, then \hl{$P'$} has the form of $C'[P]$.
Traces are used \hl{in} reasoning about the number of intermediate actions in weak
transitions. For instance, from Def.~\ref{d:BisCon}, it is easy
to see that a weak transition either becomes shorter
or remains the same when moving between $\mcontrBIS$-related processes.
This property is essential in the proof of
Lemma~\ref{l:uptocon}. We show only one such lemma, for the case of
visible weak transitions:
% passing into $\mcontrBIS$ (from left to right):
\begin{alltt}
\HOLTokenTurnstile{} \HOLFreeVar{P} \HOLSymConst{\HOLTokenContracts{}} \HOLFreeVar{Q} \HOLSymConst{\HOLTokenImp{}}
   \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{xs} \HOLBoundVar{l} \ensuremath{\HOLBoundVar{P}\sp{\prime}}.
       \HOLConst{TRACE} \HOLFreeVar{P} \HOLBoundVar{xs} \ensuremath{\HOLBoundVar{P}\sp{\prime}} \HOLSymConst{\HOLTokenConj{}} \HOLConst{UNIQUE_LABEL} \ensuremath{(}\HOLConst{label} \HOLBoundVar{l}\ensuremath{)} \HOLBoundVar{xs} \HOLSymConst{\HOLTokenImp{}}
       \HOLSymConst{\HOLTokenExists{}}\ensuremath{\HOLBoundVar{xs}\sp{\prime}} \ensuremath{\HOLBoundVar{Q}\sp{\prime}}.
           \HOLConst{TRACE} \HOLFreeVar{Q} \ensuremath{\HOLBoundVar{xs}\sp{\prime}} \ensuremath{\HOLBoundVar{Q}\sp{\prime}} \HOLSymConst{\HOLTokenConj{}} \HOLFreeVar{P} \HOLSymConst{\HOLTokenContracts{}} \HOLFreeVar{Q} \HOLSymConst{\HOLTokenConj{}} \HOLConst{LENGTH} \ensuremath{\HOLBoundVar{xs}\sp{\prime}} \HOLSymConst{\HOLTokenLeq{}} \HOLConst{LENGTH} \HOLBoundVar{xs} \HOLSymConst{\HOLTokenConj{}}
           \HOLConst{UNIQUE_LABEL} \ensuremath{(}\HOLConst{label} \HOLBoundVar{l}\ensuremath{)} \ensuremath{\HOLBoundVar{xs}\sp{\prime}}\hfill{[contracts_AND_TRACE_label]}
\end{alltt}

### 2019-11-26 ic-journal/related.tex
text

Monica Nesi did the first CCS formalisations for both pure and
value-passing CCS \cite{Nesi:1992ve,Nesi:2017wo} using early versions of the HOL
theorem prover.\footnote{Part of this work can now be found at
  \url{https://github.com/binghe/HOL-CCS/tree/master/CCS-Nesi}.}
Her main focus was on implementing decision procedures (as a ML
program, e.g.~\cite{cleaveland1993concurrency}) for
automatically proving bisimilarities of CCS processes.
Her work has been a basis for ours~\cite{Tian:2017wrba}.
However, the differences are substantial, especially in the way of defining
bisimilarities. We greatly benefited from \hl{new} features and standard
libraries in recent versions of HOL4, and our formalisation has
covered a much larger spectrum of the (pure) CCS theory.

Bengtson, Parrow and Weber did a substantial formalisation work
on CCS, $\pi$-calculi
and $\psi$-calculi 
using Isabelle/HOL and its nominal logic, with the main focus on the handling of
name binders \cite{bengtson2007completeness,parrow2009formalising}.
More details can be found in Bengtson's PhD thesis~\cite{bengtson2010formalising}. For CCS, 
he has formalized basic properties for strong/weak equivalence (congruence, basic
 algebraic laws); \hl{however the CCS syntax doesn't have constants
(or recursion operator), using replication in place of recursion.}
% The formalisation effort has then been continued 
%
% On the other side, Jesper
% Bengtson and Joachim Parrow made great progress on $\pi$-calculus
% formalization and 
% proved that the algebraic axiomatization of bisimulation
% equivalence in the $\pi$-calculi is sound and
% complete. \cite{bengtson2007completeness} 
Other formalisations in this area include the early work of T.F.~Melham
\cite{melham1994mechanized} and O.A.~Mohamed
\cite{mohamed1995mechanizing} in HOL, Compton
\cite{compton2005embedding} in Isabelle/HOL,
Solange\footnote{\url{https://github.com/coq-contribs/ccs}} in Coq
and Chaudhuri et al.\;\cite{chaudhuri2015lightweight} in Abella, the latter
focuses on `bisimulation up-to' techniques (for strong bisimilarity)
for CCS and $\pi$-calculus.
Damien Pous \cite{pous2007new} also formalised up-to techniques and some CCS examples in
Coq.
Formalisations less related to ours
include Kahsai and Miculan \cite{kahsai2008implementing} for the spi
calculus in Isabelle/HOL, and Hirschkoff \cite{hirschkoff1997full} for the $\pi$-calculus in Coq.

>>>>

Monica Nesi did the first CCS formalisations for both pure and
value-passing CCS \cite{Nesi:1992ve,Nesi:2017wo} using early versions of the HOL
theorem prover.\footnote{Part of this work can now be found at
  \url{https://github.com/binghe/HOL-CCS/tree/master/CCS-Nesi}.}
Her main focus was on implementing decision procedures (as a ML
program, e.g.~\cite{cleaveland1993concurrency}) for
automatically proving bisimilarities of CCS processes.
Her work has been a basis for ours~\cite{Tian:2017wrba}.
However, the differences are substantial, especially in the way of defining
bisimilarities. We greatly benefited from \hl{new} features and standard
libraries in recent versions of HOL4, and our formalisation has
covered a  larger spectrum of the (pure) CCS theory.

Bengtson, Parrow and Weber did a substantial formalisation work
on CCS, $\pi$-calculi
and $\psi$-calculi 
using Isabelle/HOL and its nominal logic, with the main focus on the handling of
name binders \cite{bengtson2007completeness,parrow2009formalising}.
More details can be found in Bengtson's PhD thesis~\cite{bengtson2010formalising}. For CCS, 
he has formalized basic properties for strong/weak equivalence (congruence, basic
 algebraic laws); \hl{the CCS syntax does not have constants
or recursion, using instead replication.}
% The formalisation effort has then been continued 
%
% On the other side, Jesper
% Bengtson and Joachim Parrow made great progress on $\pi$-calculus
% formalization and 
% proved that the algebraic axiomatization of bisimulation
% equivalence in the $\pi$-calculi is sound and
% complete. \cite{bengtson2007completeness} 
Other formalisations in this area include the early work of T.F.~Melham
\cite{melham1994mechanized} and O.A.~Mohamed
\cite{mohamed1995mechanizing} in HOL, Compton
\cite{compton2005embedding} in Isabelle/HOL,
Solange\footnote{\url{https://github.com/coq-contribs/ccs}} in Coq
and Chaudhuri et al.\;\cite{chaudhuri2015lightweight} in Abella, the latter
focuses on `bisimulation up-to' techniques (for strong bisimilarity)
for CCS and $\pi$-calculus.
Damien Pous \cite{pous2007new} also formalised up-to techniques and some CCS examples in
Coq.
Formalisations less related to ours
include Kahsai and Miculan \cite{kahsai2008implementing} for the spi
calculus in Isabelle/HOL, and Hirschkoff \cite{hirschkoff1997full} for the $\pi$-calculus in Coq.
